---
title: "Assignment 1 Exploratory Data Analysis"
author: "Xavier Vílchez (xavier.vilchez01@estudiant.upf.edu)"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"      
output:
  html_document:
    toc: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Analysis of the Heart Disease Dataset 
Load the data from
[here](https://raw.githubusercontent.com/jpinero/DMI_2021/main/datasets/heart_disease_dataset.csv), and the description is [here](https://raw.githubusercontent.com/jpinero/DMI_2021/main/datasets/heart_disease_description.txt). 
The original dataset comes from [here](https://archive.ics.uci.edu/ml/datasets/Heart+Disease) and corresponds to the [processed cleveland data](https://archive.ics.uci.edu/ml/machine-learning-databases/heart-disease/processed.cleveland.data)

## Perform an EDA on the dataset

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Load necessary libraries
library(mice)
library(GGally)
library(DataExplorer)
library(SmartEDA)
library(vtable)
library(mlbench)
library(ggplot2)
library(tidyr)
library(dplyr)
library(reshape2)
library(ggpubr)
library(rcompanion)
library(corrplot)
library(readxl)
library(gridExtra)
library(grid)
library(ggplotify)
library(patchwork)
library(naniar)
library(ComplexHeatmap)
library(circlize)
library(tibble)
library(viridis)

```

```{r}
# Load the dataset
data <- read.table("./data/raw_data/heart_disease_dataset.csv", na.strings = "?", header = TRUE)

# Convert appropriate columns to factors for categorical variables
factor_columns <- c("sex", "cp", "fbs", "restecg", "exang", "slope", "ca", "thal", "num")
data[factor_columns] <- lapply(data[factor_columns], as.factor)

# 1. Data summary reports
# Generate summaries using vtable
vtable(data)

#Check the quality of the data
plot_intro(data)

# 2. Check for missing values
# Visualize missing data
plot_missing(data)
```

The dataset contains clinical data from 303 patients with 15 variables related to heart disease or health. 

As we can see, we have almost no NAs in the dataset, but we will try to impute them using the Mice package.

```{r}
# Handle missing values using mice
if (anyNA(data)) {
  invisible(capture.output({
      mice_output <- mice(data, m = 1, method = 'pmm', seed = 1)
  data_imputed <- complete(mice_output, 1) # Use the first imputed dataset
    }))
}

# Check if it's okay
plot_missing(data_imputed)
```


The imputation process has been successful, now we will tackle the outlier detection.


```{r}
# Identify numerical columns, ignoring the ID column
numerical_columns <- setdiff(names(data_imputed), c("patient_id", factor_columns))

# Update the 'sex' variable to use "Male" and "Female" instead of 1 and 0
data_imputed$sex <- factor(data_imputed$sex, levels = c(0, 1), labels = c("Female", "Male"))

# 3. Outliers detection and replacement

# Detect and replace outliers
data_cleaned <- data_imputed
replace_outliers <- function(column_name) {
  column <- data_cleaned[[column_name]]
  if (is.numeric(column)) {
    Q1 <- quantile(column, 0.25)
    Q3 <- quantile(column, 0.75)
    IQR <- Q3 - Q1 
    upper_limit <- Q3 + 1.5 * IQR 
    lower_limit <- Q1 - 1.5 * IQR 
    data_cleaned[[column_name]] <<- ifelse(column > upper_limit, upper_limit, ifelse(column < lower_limit, lower_limit, column))
  }
}

# Box plots for numerical variables
# Apply the function to the numerical columns
invisible(lapply(numerical_columns, replace_outliers))

# Function to create comparison box plots before and after outliers replacement 
create_comparison_boxplots <-function(data_before, data_after, numerical_columns) {
  #Create a list to store all plots 
  plot_list <- list()
  
  for (col in numerical_columns) {
    # Prepare data for plotting 
    df_combined <- data.frame(
      value = c(data_before[[col]], data_after[[col]]), 
      group = factor(rep(c("Before", "After"), each = nrow(data_before)), 
                     levels = c("Before", "After")),
      variable = col
    )
    # Create plot
    p <- ggplot(df_combined, aes(x = group, y = value, fill = group, color = group)) +
      # Add boxplot
      geom_boxplot(alpha = 0.5, outlier.shape = NA) +
      # Add jittered points 
      geom_jitter(width = 0.2, alpha = 0.3) +
      # Set colors 
      scale_fill_manual(values = c("Before" = "purple", "After" = "darkblue")) +
      scale_color_manual(values = c("Before" = "purple", "After" = "darkblue")) +
      # Customize theme 
      theme_minimal() +
      theme(
        plot.title = element_text(size = 12, face = "bold"), 
        axis.title.x = element_blank(),
        panel.grid.major.x = element_blank()
      ) +
      # Add labels 
      labs(title = paste("Distribution of", col), y = col) + xlab(col)
    plot_list[[col]] <- p
  }
  return(plot_list)
}

#Create all plots 
plots <- create_comparison_boxplots(data_imputed, data_cleaned, numerical_columns)
plots 
```


As you can see in the boxplots, the outliers have been replaced by values 3SD from the mean. Now that we have a clean dataset, we can create a file for future use.

```{r}
# Save the processed and imputed dataset
write.csv(data_cleaned, "./data/processed_data/heart_disease_dataset_cleaned.csv", row.names = FALSE)
```


With the clean dataset, we can now perform an EDA, starting with the univariate analysis for continuous and categorical variables.

```{r}
# 4. Univariate summaries

summary(data_cleaned[numerical_columns])

# Function to create a combined frequency table for all categorical variables 
create_combined_freq_table <- function(data, factor_columns) {
  # Initialize an empty data frame to store all results 
  all_frequencies <- list()
  
  # Process each factor column in order 
  for (col in factor_columns) {
    # Calculate frequencies 
    freq <- table(data[[col]])
    rel_freq <- prop.table(freq) * 100
    
    #Create data frame for this variable 
    freq_df <- data.frame(
      Variable = col, 
      Category = names(freq), 
      Frequency = as.numeric(freq), 
      Percentage = round(as.numeric(rel_freq), 2)
    )
    # Append to the main data frame 
    all_frequencies[[col]] <- freq_df
  }
  all_frequencies <- do.call(rbind, all_frequencies)
  
  # Create nice looking table with kableExtra
  kable(all_frequencies, 
        format = "html", 
        caption = "Frequency distribution of categorical variables") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
}

# Create and display the combined frequency table 
combined_freq_table <- create_combined_freq_table(data_cleaned, factor_columns)
combined_freq_table

# Plot univariate distributions
plot_bar(data_cleaned[factor_columns], title = "Bar Plots for Categorical Variables")

plot_histogram(data[ , numerical_columns], title = "Histograms for Numerical Variables")
```


From the barplots we can see that the sex ratio is imbalanced in this study, as for the rest of the variables, we see that most of the patients have low levels of the possible biomarkers for having heart disease and that they actually doesn't show any symptoms related with heart disease (represented by the increase in the factor num).

From the histograms we can see that the age follows approximately a normal distribution with mean 54 years old (younger people typically don't have this kind of disease). Chol, oldpeak and trestbps have left-skewed distributions, whereas thalach follows a right-skewed distribution. However, the median of the Chol variable is at 241, which by the medical standards is considered high.

Now, we will proceed with the bivariate analysis.

```{r, warning=FALSE}
# 5. Bivariate analysis
# Pairwise scatter plots and correlation for numerical variables
ggpairs(data_cleaned, columns = c("age", "trestbps", "chol", "thalach", "oldpeak"))

# Function to compute Cramér's V
cramers_v <- function(x, y) {
  tbl <- table(x, y)
  chisq <- chisq.test(tbl)
  n <- sum(tbl)
  phi2 <- chisq$statistic / n
  k <- min(ncol(tbl), nrow(tbl))
  return(sqrt(phi2 / (k - 1)))
}

# Compute Cramér's V for all categorical variable pairs
factor_vars <- factor_columns[factor_columns != "num"]
cramers_matrix <- matrix(NA, nrow = length(factor_vars), ncol = length(factor_vars),
                         dimnames = list(factor_vars, factor_vars))

for (i in 1:length(factor_vars)) {
  for (j in 1:length(factor_vars)) {
    if (i != j) {
      cramers_matrix[i, j] <- cramers_v(data_cleaned[[factor_vars[i]]], data_cleaned[[factor_vars[j]]])
    } else {
      cramers_matrix[i, j] <- NA  # Hide diagonal
    }
  }
}

# Convert to long format for ggplot
cramers_df <- melt(cramers_matrix, na.rm = TRUE)  # Remove NA values (diagonal)

# Plot heatmap without diagonal
ggplot(cramers_df, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "red", na.value = "white") +
  labs(title = "Association Between Categorical Variables (Cramér's V)",
       x = "Variable 1", y = "Variable 2", fill = "Cramér's V") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


From the bivariate continuous analysis, we can see that age has significant correlation with all the other variables. 

As for the categorical variables, we have used Cramer's V test to see which variables are more related to each other, and we can see that the most related pairs are sex-thal and exang-cp.

## Create visualizations in order to show which variables seem to be more associated with heart disease

```{r, warning=FALSE}
# 7. Association between variables and heart disease
# Chi-squared test and filtering significant variables
significant_categorical_vars <- c()

for (col in setdiff(factor_columns, "num")) {
  chisq_test <- chisq.test(table(data_cleaned[[col]], data_cleaned$num))
  p_value <- chisq_test$p.value
  
  if (p_value < 0.05) {
    significant_categorical_vars <- c(significant_categorical_vars, col)
  }
}

# Ensure 'num' is a factor for proper grouping
data_cleaned$num <- as.factor(data_cleaned$num)

# Plot only significant categorical variables
for (col in significant_categorical_vars) {
  p <- ggplot(data_cleaned, aes(x = !!sym(col), fill = num)) +
    geom_bar(position = "fill") +
    labs(title = paste("Proportion of Heart Disease by", col),
         x = col, y = "Proportion") +
    theme_minimal() +
    scale_fill_brewer(palette = "Set3")
  print(p)
}
```


The plots from the significant Chi-squared results indicate that all the significant categorical variables associated with num (which is the factor describing the severity of the heart disease) show a positive linear correlation with num.

```{r}
# Calculate correlations and their p-values for numerical variables vs. heart disease

correlation_results <- data.frame(Variable = character(), Correlation = numeric(), P_Value = numeric())

for (col in numerical_columns) {
  if (is.numeric(data_cleaned[[col]])) {
    cor_test <- cor.test(data_cleaned[[col]], as.numeric(data_cleaned$num), method = "spearman", exact = FALSE)
    correlation_results <- rbind(correlation_results,
                                  data.frame(Variable = col,
                                             Correlation = cor_test$estimate,
                                             P_Value = cor_test$p.value))
  }
}


# Visualize the correlation coefficients and their significance
ggplot(correlation_results, aes(x = reorder(Variable, Correlation), y = Correlation, fill = P_Value < 0.05)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("red", "green"), labels = c("Not Significant", "Significant")) +
  labs(title = "Correlation of Numerical Variables with Heart Disease",
       x = "Numerical Variable",
       y = "Spearman Correlation Coefficient") +
  coord_flip() +
  theme_minimal()

# Store correlation results in a dataframe efficiently
correlation_results <- data.frame(Variable = character(), Correlation = numeric(), P_Value = numeric())

for (col in numerical_columns) {
  if (is.numeric(data_cleaned[[col]])) {
    cor_test <- cor.test(data_cleaned[[col]], as.numeric(data_cleaned$num), method = "spearman", exact = FALSE)
    correlation_results <- rbind(correlation_results,
                                  data.frame(Variable = col,
                                             Correlation = cor_test$estimate,
                                             P_Value = cor_test$p.value))
  }
}

# Filter only significant variables (p < 0.05)
significant_vars <- correlation_results %>%
  filter(P_Value < 0.05) %>%
  pull(Variable)

# Ensure heart disease variable "num" is included
significant_vars <- c(significant_vars, "num")

# Ensure 'num' is numeric
data_cleaned$num <- as.numeric(as.character(data_cleaned$num)) 

# Compute correlation matrix only for significant numerical variables
cor_matrix_filtered <- cor(data_cleaned[significant_vars], method = "spearman", use = "pairwise.complete.obs")

# Visualize the filtered correlation matrix
corrplot(cor_matrix_filtered, method = "color", type = "upper", 
         tl.col = "black", tl.srt = 45, # Rotate labels for better readability
         addCoef.col = "black", # Show correlation coefficient values
         diag = FALSE) # Hide diagonal (correlation of variables with themselves)
```


As for the continuous variables, the correlation matrix obtained with the significant Spearman's correlation results shows that oldpeak is the most positively associated variable with num, and that a increase in the thalach variable decreases the severity of the disease.

# 2 Difference in mortality rates in hospitalized COVID-19 patients 
*Using the supplementary material from the [Difference in mortality rates in hospitalized COVID-19 patients identified by cytokine profile clustering using a machine learning approach: An outcome prediction alternative](https://www.frontiersin.org/articles/10.3389/fmed.2022.987182/full), perform the following tasks*

## Reproduce Figure 1 from the publication

```{r, message=FALSE, warning=FALSE}
# Load the datasets
file1 <- "./data/raw_data/Table_1_COVID-19.XLSX"

data1 <- read_excel(file1)

# Replace the headers of data1 with the first row
colnames(data1) <- data1[1, ]  # Set the first row as column names
data1 <- data1[-1, ]           # Remove the first row

# Replace "NI" and "ND" with NA in all columns of data1
data1[data1 == "NI" | data1 == "ND"] <- NA

# Remove exact duplicates
data1 <- data1 %>%
  distinct()

# Remove duplicated IDs with different data
data1 <- data1 %>%
  group_by(ID) %>%
  filter(n() == 1) %>%
  ungroup()

# Remove rows with the Death value being different than Yes or No
data1 <- data1 %>%
  filter(!is.numeric(as.numeric(Death)) | Death %in% c("Yes", "No"))

# Iterate over each row to apply the conditions
data1$`ARDS Diagnosis` <- ifelse(
  !(data1$`ARDS Diagnosis` %in% c("Yes", "No")),  # Check if ARDS Diagnosis is not "Yes" or "No"
  ifelse(data1$`Days with ARDS` == 0, "No", "Yes"),  # If Use of ARDS is 0, set Diagnosis to "No", else "Yes"
  data1$`ARDS Diagnosis`  # Keep the existing value if ARDS Diagnosis is already "Yes" or "No"
)

data1$`Use of NIV` <- ifelse(
  !(data1$`Use of NIV` %in% c("Yes", "No")),  # Check if ARDS Diagnosis is not "Yes" or "No"
  ifelse(data1$`Days with NIV` == 0, "No", "Yes"),  # If Use of ARDS is 0, set Diagnosis to "No", else "Yes"
  data1$`Use of NIV`  # Keep the existing value if ARDS Diagnosis is already "Yes" or "No"
)

# Create the clinical groups
# Example: Assign groups based on a column (e.g., "age")
# Add clinical_class column based on conditions
data1 <- data1 %>%
  mutate(clinical_class = case_when(
    `Use of NIV` == "No" & `Use of AMV` == "No" & `ARDS Diagnosis` == "No" ~ "G1",
    `Use of NIV` == "Yes" & `Use of AMV` == "No" & `ARDS Diagnosis` == "Yes" ~ "G3",
    `Use of AMV` == "Yes" & `ARDS Diagnosis` == "Yes" ~ "G4",
    TRUE ~ "G2"  # Default case for unmatched rows
  ))

# Save processed data
write.csv(data1, "./data/processed_data/Table_1_COVID-19_processed.csv", row.names = FALSE)
```

This dataset contains clinical and cytokine data from 199 hospitalized COVID-19 patients with 19 columns. 

```{r}
# --- Reproduce the Figure ---

#### Plot 1: Age Histogram (Top-Left)
data1$Age <- as.numeric(data1$Age)
age_histogram <- hist(data1$Age,
         breaks = seq(20, 100, by = 10),
         col = "#e0eeee",
         border = "black",
         main = "Age",
         xlab = "Age (years)",
         ylab = "Frequency (n)",
         ylim = c(0, 50),
         axes = TRUE)

#### Plot 2: Clinical Classification Table (Top-Right)
# Create the table with 4 columns and 4 rows
table_1B <- data.frame(matrix(ncol = 4, nrow = 4))

# Assign the column names to the table
colnames(table_1B) <- c("Clinical\nclassification", "NIV", "AMV", "ARDS")

# Set the corresponding values
table_1B[1,] <- c("G1", "-", "-", "-")
table_1B[2,] <- c("G2", "-/+", "+/-", "-")
table_1B[3,] <- c("G3", "+", "-", "+")
table_1B[4,] <- c("G4", "-/+", "+", "+")

# Create table with improved formatting
table_1B <- tableGrob(table_1B, 
                     rows = NULL,
                     theme = ttheme_default(
                         core = list(fg_params=list(cex = 0.8)),
                         colhead = list(fg_params=list(cex = 0.8))))
grid::grid.newpage()  # Start a new plotting page
grid::grid.draw(table_1B)  # Draw the table
grid::grid.text("Definition of the clinical classification", y = 0.95, gp = gpar(fontsize = 14, fontface = "bold"))

#### Plot 3: Clinical class barplot (Bottom-Left)
clinical_class_plot <-mids <- barplot(table(data1$clinical_class),
                    col = c("#8dd3c8", "#feffb3", "#bebbda", "#fb8071"),
                    main = "Clinical class",
                    xlab = "Clinical classification",
                    ylab = "Frequency (n)",
                    ylim = c(0, 90))  
    
    text(x = mids, 
         y = table(data1$clinical_class) + 5, 
         labels = table(data1$clinical_class), 
         cex = 1.2)

#### Plot 4: Vital Status barplot (Bottom-Right)
death_plot <-mids <- barplot(table(data1$Death),
                    col = c("#8dd3c8", "#feffb3"),
                    main = "Vital Status",
                    xlab = "Death",
                    ylab = "Frequency (n)",
                    ylim = c(0, max(table(data1$Death)) + 40))  
    
    text(x = mids,
         y = table(data1$Death) + 10,
         labels = table(data1$Death),
         cex = 1.2)
    
### We tried to put the plots together using these and transforming the rbase plots to ggplots (with as.ggplot function), but it didn't work
    
# wrap_elements(panel = age_histogram, clip = FALSE) + 
#   table_1B +
#   wrap_elements(panel = clinical_class_plot, clip = FALSE) +
#   wrap_elements(panel = death_plot, clip = FALSE) + 
#   plot_annotation(tag_levels = 'A') # Tag plots

```


## Reproduce Figure 2 from the publication
*but instead of representing the clusters in the annotation, represent the groups (G1 to G4)*

```{r, warning=FALSE, message=FALSE}

# Data import and initial processing 
file2 <- "./data/raw_data/Table_2_COVID-19.XLSX"
data2 <- read_excel(file2)

# Clean column headers 
colnames(data2)[-1] <- data2[1, -1]  # Replace all but the first column
data2 <- data2[-1, ]                # Remove the first row

# Fill missing values in ID column
data2 <- data2 %>%
  fill(ID, .direction = "down")

# Replace "NI" with NA 
data2[data2 == "NI"] <- NA

# Remove unnecessary columns 
data2 <- data2[ ,-2] #Remove Day column
data2 <- data2[, -((ncol(data2)-12):ncol(data2))]

# Convert columns to numeric 
data2 <- data2 %>%
  mutate(across(-ID, ~ as.numeric(.), .names = "{col}"))

# Process duplicated IDs  
data2_processed <- data2 %>%
  group_by(ID) %>%
  summarise(across(everything(), ~ mean(.x, na.rm = TRUE))) %>%
  mutate(across(-ID, ~ round(., 4)))

# Replace NaN with NA
data2_processed[is.nan(data2_processed)] <- NA

# View missing values 
vis_miss(data2_processed)

# Impute missing values 
imputation <- mice(data2_processed, method = "pmm", m = 5)
data2_processed <- complete(imputation)

# Check there are any missing values
vis_miss(data2_processed)

# Fix ID mistakes 
data2_processed$ID <- gsub("COAG-HSDJ-", "COAG-HSJD-", data2_processed$ID)

# Merge with clinical classes 
data2_processed <- merge(data2_processed, data1[, c("ID", "clinical_class")], by = "ID", all.x = TRUE)

# Save processed data
write.csv(data2_processed, "./data/processed_data/Table_2_COVID-19_processed.csv", row.names = FALSE)
```

To reproduce the Figure 2 of the publication, first we processed the data. Once processed, the data contains 194 COVID-19 patients with 22 variables related with the disease. 
As we can see in the plots, we had missing values in the column of IL-6. But, then the imputation was successful. 

```{r}
# ---- Reproduce Heatmap Figure ---

# Select the cytokines and convert to numeric
cytokines <- data2_processed[, c("IL-1β", "IL-6",  "IL-10", "IFN-ɑ", "TNF-ɑ", "IL-8", "G-CSF", "IFN-γ", "CCL3", "CXCL10", "CCL2", "IL-38")]
cytokines <- as.data.frame(lapply(cytokines, as.numeric))

# Process row annotations 
row_ann <- data2_processed$clinical_class
names(row_ann) <- data2_processed$ID

# Remove rows with missing annotations 
valid_rows <- !is.na(row_ann)
cytokines_clean <- cytokines[valid_rows, ]
row_ann_clean <- row_ann[valid_rows]

#Normalize data (0-100 scale)
cytokines_clean <- as.data.frame(lapply(cytokines_clean, function(x) {
  x <- x / max(x, na.rm = TRUE) * 100
}))

#Create annotation dataframe
ann_df <- data.frame(ClinicalClass = row_ann_clean)
rownames(cytokines_clean) <- rownames(ann_df)

# Prepare matrix for heatmap
cytokines_clean <- t(cytokines_clean)
ann_df_order <- ann_df[order(ann_df$ClinicalClass), , drop = FALSE]
cytokines_clean_order <- cytokines_clean[, rownames(ann_df_order)]

#Fix cytokine names 
rownames(cytokines_clean_order) <- gsub("\\.", "-", rownames(cytokines_clean_order))

#Define annotation colors 
group_colors <- list(ClinicalClass = c(
  "G1" = "#0fbc40",
  "G2" = "#629cff",
  "G3" = "#f87772",
  "G4" = "#cd0bbc"
))

#Create top annotation
top_anno <- HeatmapAnnotation(
  ClinicalClass = ann_df_order$ClinicalClass, 
  col = group_colors,
  show_annotation_name = FALSE, 
  annotation_legend_param = list(
    ClinicalClass = list(
      title = "Clinical class", 
      at = c("G1", "G2", "G3", "G4")
    )
))

# We tried to have the same order as in the original Figure, but it didn't work. 
# cyt_order <- c("IL.6", "CXCL10", "IL.38", "IL.8", "IFN.ɑ", "IL.10", "TNF.ɑ", "CCL2", "CCL3", "IFN.γ", "IL.1β", "G.CSF")
# cytokines_clean_order <- cytokines_clean_order[cyt_order, ]

#Create heatmap
heatmap1 <- Heatmap(cytokines_clean_order, 
        name = "Relative Expression", 
        col = colorRamp2(c(0, 50, 100), colors = c("#fdeee1", "#fc9a55", "#8c2c06")), 
        cluster_rows = TRUE, 
        cluster_columns = TRUE, 
        top_annotation = top_anno, 
        show_row_names = TRUE, 
        show_column_names = FALSE, 
        row_names_side = "right",
        column_title = "Patients", 
        column_title_side = "bottom", 
        heatmap_legend_param = list(
           title = "Relative Expression (%)", 
           at = c(0, 50, 100), 
           labels = c("0.00", "50.0½", "100.0%"),
           color_bar = "discrete"
           )
)

# Draw final heatmap
draw(heatmap1, 
  heatmap_legend_side = "left", 
  annotation_legend_side = "left")

```




## Improve figure 2 of the publication
*Add a second annotation with information of deaths and a third one with information of gender*

```{r}

# Merge the data with metadata
data2_processed_meta <- merge(data2_processed, data1[, c("ID", "Gender", "Death")], by = "ID", all.x = TRUE)

# Check for missing data
vis_miss(data2_processed_meta)

# Filter out rows with NA in key columns
data2_processed_meta <- data2_processed_meta %>%
  filter(!is.na(Gender) & !is.na(Death) & !is.na(clinical_class))

# Verify no missing data remains
vis_miss(data2_processed_meta)

# Remove the specific row where Gender is 72 
data2_processed_meta <- data2_processed_meta[data2_processed_meta$Gender %in% c("M", "F"), ]

# Save merged data with processed data plus metadata 
write.csv(data2_processed_meta, "./data/processed_data/Table_2_COVID-19_processed_with_metadata.csv", row.names = FALSE)
```


To improve the Figure 2 of the publication, after merging the processed dataset 2 with the metadata, we checked the missing values and imputed them. 

```{r}
# Select cytokine columns and convert to numeric
cytokines2 <- data2_processed_meta[, c("IL-1β", "IL-6", "IL-10", "IFN-ɑ", 
                                      "TNF-ɑ", "IL-8", "G-CSF", "IFN-γ", 
                                      "CCL3", "CXCL10", "CCL2", "IL-38")]
cytokines2 <- as.data.frame(lapply(cytokines2, as.numeric))

# Create matrix for heatmap
cytokine2_matrix <- as.matrix(cytokines2)
rownames(cytokine2_matrix) <- data2_processed_meta$ID

# Normalize the data (0-100 scale)
cytokine2_matrix_norm <- apply(cytokine2_matrix, 2, function(x) {
  (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE)) * 100
})

# Create a new ordered factor for clinical_class to ensure specific order
data2_processed_meta$clinical_class <- factor(data2_processed_meta$clinical_class, levels = c("G1", "G2", "G3", "G4"))

# Create column annotation 
col_anno2 <- HeatmapAnnotation(
  Gender = factor(data2_processed_meta$Gender), 
  Death = factor(data2_processed_meta$Death), 
  ClinicalClass = factor(data2_processed_meta$clinical_class),
  col = list(
    Gender = c("F" = "pink", "M" = "skyblue"), 
    Death = c( "No" = "white", "Yes" = "black"), 
    ClinicalClass = c(
    "G1" = viridis(4)[1],
    "G2" = viridis(4)[2],
    "G3" = viridis(4)[3],
    "G4" = viridis(4)[4]
  )), 
  show_annotation_name = FALSE, 
  annotation_legend_param = list(
    ClinicalClass = list(title = "Clinical class")
  )
)

# Create and display the heatmap
heatmap2 <- Heatmap(
    t(cytokine2_matrix_norm),
    name = "Expression (%)",
    col = colorRamp2(c(0, 50, 100), c("#1F449C", "white", "#F05039")),
    top_annotation = col_anno2, 
    show_row_names = TRUE,
    show_column_names = FALSE,
    cluster_rows = TRUE,
    cluster_columns = FALSE,
    column_title = NULL, 
    column_order = order(
      data2_processed_meta$clinical_class, 
      data2_processed_meta$Gender, 
      data2_processed_meta$Death
    ),
    heatmap_legend_param = list(
        at = c(0, 50, 100),
        labels = c("0", "50", "100"), 
        title = "Expression (%)"
    )
)

# Draw the heatmap
draw(heatmap2)
```


The heatmap displays cytokine expression levels across different samples, using a color-blind-friendly palette. We can see a clear grouping based on clinical class (G1-G4). The top annotations also represent gender and survival status. 


# session info {.unnumbered}

```{r, results='asis',  echo=FALSE, message=FALSE }
sessionInfo()
```
